# Spatial coverage sampling {#SpatialCoverage}

Local undersampling with regular grids can be avoided by relaxing the constraint that the sampling locations are restricted to the nodes of a regular grid. This is what is done in spatial coverage sampling or, in case of a sample that is added to an existing sample, in spatial *infill* sampling. Spatial coverage and infill samples cover the area or fill in the empty space as uniformly as possible. The sampling locations are obtained by minimizing a criterion that is defined in terms of the geograhic distances between the nodes of a fine interpolation grid and the sampling locations. @bru07c proposed to minimize the mean of the squared distances of the grid nodes to their nearest sampling location (Mean Squared Shortest Distance):
\begin{equation}
MSSD=\frac{1}{N}\sum_{i=1}^{N}\min_{j}\left(D_{ij}^{2}\right) \;,
(\#eq:MSSD)
\end{equation}
where $N$ is the total number of nodes of the interpolation grid, and $D_{ij}$ is the distance between the $i$th grid node and the $j$th sampling location. This distance measure can be minimized by the k-means algorithm. The same algorithm is used in Chapter \@ref(STSI) to construct compact geographical strata (shortly referred to as geostrata) for stratified random sampling. In stratified random sampling, one or more sampling locations are selected randomly from each geostratum. For mapping purposes probability sampling is not required. For each geostratum the means of the spatial coordinates of the nodes of that geostratum are calculated, and these centroids are used as sampling locations. This improves the spatial coverage compared to stratified *random* sampling.

In probability sampling we may want to have strata of equal area (clusters with equal numbers of nodes), so that the sampling design becomes self-weighting. For mapping this constraint is not recommended as it may lead to samples with suboptimal spatial coverage. Existing sampling locations can easily be accommodated in the k-means algorithm, by using them as fixed centroids. Spatial coverage and infill samples can be computed with ```R``` package ```spcosa``` [@walvoort2010]. Figure \@ref(fig:spatcovVoorst) shows a spatial coverage sample of the same size as the regular grid in study area Voorst (Figure \@ref(fig:gridVoorst)). Note that the undersampled area in the centre of the study area is now covered by a sampling location.

```{r spatcovVoorst, out.width='100%',fig.asp=0.333, fig.cap="Spatial coverage sample from Voorst"}
#set number of sampling locations to be selected
n<-115
set.seed(314)
#compute compact geostrata
gridded(grdVoorst) <- ~s1+s2
myStrata <- stratify(grdVoorst, nStrata = n, equalArea=FALSE, nTry=10)
#select the centres of the geostrata
mySample <- spsample(myStrata)
#plot geostrata and sampling points (centres of geostrata)
plot(myStrata, mySample)
```

## Spatial infill sampling
If georeferenced data are available that can be used for mapping the variable of interest, but we need more data for mapping, it is attractive to account for these existing sampling locations when selecting the additional sampling locations. The aim now is to fill in the empty spaces, i.e. the parts of the study area not covered by the existing sampling locations. This is referred to as spatial infill sampling. Figure \@ref(fig:spatialinfillEthiopia) shows a spatial infill sample in three woredas in Ethiopia. The existing data are mainly along roads. Besides, one woreda, the one in tht northeast, is already covered quite well by the existing data.

```{r spatialinfillEthiopia, out.width='70%',fig.asp=1, fig.cap="Spatial infill sample in three woredas of Ethiopia", message=F, cache=T}
gridded(grdEthiopia)<-~s1+s2
#Set number of new sampling locations to be selected
n<-100
#Compute total sample size (existing points + new points)
ntot<-n+length(priordataEthiopia)
#Change class of d (existing points) from SpatialPointsDataFrame to SpatialPoints
priordataEthiopia<-as(priordataEthiopia,"SpatialPoints")
#grdEthiopia does not have projection attributes, whereas priordataEthiopia does. Remove projection attributes of priordataEthiopia
proj4string(priordataEthiopia)<- NA_character_
#Compute geostrata with option priorPoints
set.seed(314)
myStrata <- stratify(grdEthiopia, nStrata = ntot, priorPoints=priordataEthiopia, nTry=10)
#Select sampling points of infill sample (centres of geostrata)
mySample <- spsample(myStrata)
#Plot geostrata and sampling points (centres of geostrata)
plot(myStrata, mySample)
#Select the new points from mySample
ids <- which(mySample@isPriorPoint==F)
#Change class of mySample to data.frame
mySample <- as(mySample,"data.frame")
mySamplenew <- mySample[ids,]
```
